---
title: "P8105_hw2_qc2385"
author: "Qixuan Cao"
date: "2026-10-06"
output: github_document
---

```{r}
library(tidyverse)
```

## problem 1

access the data:
```{r load the NYC subway data}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

short paragraph about this dataset:

The cleaned NYC subway dataset contains `r nrow(subway_df)` observations and `r ncol(subway_df)` variables. 

The dataset includes subway line and station name, station latitude and longitude, routes served, entry status, vending availability, entrance type, and ADA compliance. 

These data are not completely tidy because the subway routes are stored across multiple columns (`route1` through `route11`) rather than in a single route variable.

### problem 1a
How many distinct stations are there?
```{r 1a}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```
There are 465 distinct subway stations.

### problem 1b
How many stations are ADA compliant?
```{r 1b}
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
There are 84 ADA-compliant stations.

### problem 1c
What proportion of station entrances / exits without vending allow entrance?
```{r 1c}
subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_entry = mean(entry, na.rm = TRUE)
  )
```
Among station entrances and exits without vending, the proportion that allow entrance is 37.7%.

## problem 1 reformat data
```{r subway reformat data}
subway_tidy =
  subway_df |>
  mutate(across(route8:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  ) |>
  filter(!is.na(route_name))
```
### problem 1 reformat a
How many distinct stations are there?
```{r 1 reformat a}
subway_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```
There are 60 distinct subway stations.

### problem 1 reformat b
How many stations are ADA compliant?
```{r 1 reformat b}
subway_tidy |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
There are 17 ADA-compliant stations.


## Problem 2
```{r}
library(readxl)
```

```{r clean mr.Trash Wheel}
mr_trash =
 read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N709"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel"
  )
```

```{r professer_trash}
professor_trash =
   read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M134"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.numeric(year),
    trash_wheel = "Professor Trash Wheel"
  )
```

```{r gwynnda_trash}
gwynnda_trash =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L351"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.numeric(year),
    trash_wheel = "Gwynnda"
  )
```

```{r merge}
trash_wheel_df =
  bind_rows(
    mr_trash,
    professor_trash,
    gwynnda_trash
  )
```

What was the total weight of trash collected by Professor Trash Wheel?
```{r}
trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )
```
The total weight of trash collected by Professor Trash Wheel was 282 tons.

What was the total number of cigarette butts collected by Gwynnda in June of 2022?
```{r}
trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )
```
Gwynnda collected 18120 cigarette butts in June 2022.
